Dark matter halo shapes at $z=0$ in the Auriga simulations

### 2019-08-03 Document/paper.tex
triaxiality

\begin{figure*}
\begin{center}
\includegraphics[width=0.8\columnwidth]{Lvl_4_Triax_Plane_DM.pdf}
\includegraphics[width=0.8\columnwidth]{Lvl_4_Triax_Plane_MHD.pdf}
\end{center}
\caption{Axial ratios for all simulated halos.
  Left/right panel corresponds to DMO/MHD simulations, respectively.
  Triangles (squares) represent the measurements at $R_{200}/16$
  ($R_{200}$) that correspond to physical distances of $\approx14$ kpc
  ($\approx 230$ kpc).
  Here we visualize three main population trends.
  First, in DMO simulations halos are rounder in the outskirts
  than in the inner part.
  Second, halos in MHD are rounder in the inner regions than in
  the outskirts (opposite to the DMO trend). 
  Third, halos in MHD are rounder than its DMO counterparts.
  These three points are further clarified in Figure 
  The first and second points are claryfied in Figure
  \ref{fig:triaxial_cumulative} through cumulative triaxility
  distributions}
  \label{fig:triaxiality_plane}
\end{figure*}

\begin{figure*}
\includegraphics[width=0.32\textwidth]{triaxialiy_distro_DM.pdf}
\includegraphics[width=0.32\textwidth]{triaxialiy_distro_MHD.pdf}
\includegraphics[width=0.32\textwidth]{delta_triaxiality_distro.pdf}
\caption{Triaxility cumulative distribution at five different radii.
  Right/middle panel corresponds to the cumulative distribution of
  triaxiality in DMO/MHD simulations, respectively. 
  In DMO simulations the median triaxiality at all radii is larger
  than $2/3$. Furthermore, the triaxility increases as one moves
  towards the inner part of the halo.
  In MHD simulations these trend reverses.
  The median triaxility at all radii is smaller than $1/3$ and the
  halo is less triaxial as on moves towards the stellar disc.
  The right panel, shows the cumulative distribution for the changes
  in triaxility in simulations with different physics, $\Delta
  T=T_{\rm MHD}-T_{\rm DMO}$. 
  At the virial radius all halos become less triaxial in the MHD
  simulations, the change is stronger towards the halo center.
  Only two haloes in the sample show the opposite trend of becoming
  more triaxial (but only at the smaller radii).} 
\label{fig:triaxial_cumulative}
\end{figure*}

@@

\begin{figure*}
\begin{center}
\includegraphics[width=1.0\textwidth]{correlations_twisting_triaxiality_MHD.pdf}
\end{center}
\caption{Twisting in the halo-disc alignment (measured as the standard deviation of the alignment angles)
  as a function of the halo triaxialiy at the radius indicated
  in each panel's title.
  The label with the $\rho$ value corresponds to the Spearmans rank
  correlation coefficient (mean value and uncertainty estimated via
  jackknife resampling) while the line shows the result of the
  minimum squares fit to the data.
  Twisting does not correlate with triaxility in the same way at every
  radii. The strongest correlation appears around $0.12R_{200}$ ($28\pm2$ kpc).} 
\label{fig:alignment_correlations}
\end{figure*}

\begin{figure*}
\begin{center}
\includegraphics[width=0.8\textwidth]{correlation_T_MHD_disc.pdf}
\end{center}
\caption{Correlations between the halo triaxility at different radii
  and baryonic disc properties. 
  The label with the $\rho$ value corresponds to the Spearmans rank
  correlation coefficient (mean value and uncertainty estimated via
  jackknife resampling).
  The line is the best minimum squares fit to a line.
  The x-axis in the first column is the gas density at the center of
  the galaxy with in a sphere of radius  $1$ kpc \citep{Pakmor17};
  the second column shows the disc to total mass ratio and the last
  column includes the disc optical radius defined to be the radius at which the
  $B$-band surface brightness drops below 25 mag arcsec$^{-2}$ \citep{auriga}.
  The largest correlations are found for the two smaller radii
  ($R_{200}/16$ and $R_{200}/8$).
  Extended and massive stellar discs with a low gas content at its
  core are correlated with low dark matter triaxilities.
  The correlation decreases as one approaches larger radii.}
\label{fig:disc_correlations}
\end{figure*}

\begin{figure*}
\begin{center}
\includegraphics[width=0.9\textwidth]{correlations_angles_alignment_MHD.pdf}
\end{center}
\caption{Twisting in the halo-disc alingment as a function of the same
  baryonic disc properties in Figure \ref{fig:disc_correlations}.
  The label with the $\rho$ value corresponds to the Spearmans rank
  correlation coefficient (mean value and uncertainty estimated via
  jackknife resampling).
  The line is the best minimum squares fit to a line.
  In this case the disc to total mass ratio does not correlate with the alignment twisting
  The gas density at the center and the disc optical radius still show
  a strong correlation, albeit weaker than what they had with the triaxility.}
\label{fig:twisting_correlations}
\end{figure*}

@@

Figure \ref{fig:resolution} shows the results for the six halos
available both at Level3 and Level4 resolutions.
The dotted lines compare the DMO simulations showing that down to
scales of $1$ kpc the two resolutions yield comparable results
\citep{VeraCiro11}. 
In MHD simulations the situation is different, as expected. 
The shape ratios $b/a$ and $c/a$ start to differentiate
from each other at radii smaller than $15$ kpc, tith the exception of
Au-27 that shows a good agreement in the shape measuremnts at all radii, 
Above $15$ kpc, the differences between the triaxility values at
Level4 and Level3 in MHD simulations are small and never exceed $0.1$. 
For this reason, in order to make the DM halo shape results comparable
in DMO and MHD simulations we only report measurements above $15$
kpc.}

@@

{\bf The main measurements are summarized in Table \ref{table:DMO} for the
DMO simulations and Table \ref{table:MHD} for the MHD simulations.
These results are computed at five different radii $R_{200}/16$,
$R_{200}/8$, $R_{200}/4$, $R_{200}/2$ and $R_{200}$.
In the following subsections we present and comment in greater detail
the results for the triaxility, the halo-disc alignment and the
correlations with baryonic disc properties.}

\subsection{Triaxility}

@@

In Figure \ref{fig:triaxial_cumulative} we summarize the results at
different radii using the cumulative distributions for the 
triaxility parameter $T$ defined as

@@

The left panel of Figure \ref{fig:triaxial_cumulative} shows that in
the DMO sample the triaxility has a median larger than $2/3$ (a
typical value that marks the transition from triaxiality to
prolateness) at all radii. 
Furthermore this median value increases as we move towards the inner part of the
halo.
The right panel shows how the correlations in the MHD simulations go
in the opposite direction with respect to DMO result.s
There the median triaxility is always smaller than $1/3$ (a typical
values that marks the transition from triaxility to prolateness) and this
median value decreases as we move towards the inner part of the halo.

We now quantify this change in triaxility at the level of individual halos.
We compute $\Delta T\equiv T_{\rm MHD}-T_{\rm DMO}$ to quantify the
change between the triaxility in the MHD and the DMO simulation.
Figure \ref{fig:delta_triaxial_cumulative} shows the cumulative
distribution at the same radii as in
Figure \ref{fig:delta_triaxial_cumulative}.  
From this Figure we observe that at $R_{200}$ all the halos have
reduced its triaxility include baryonic physics.
At smaller radii this general trend continues, although 2 haloes (Au-11 and Au-17) 
in our sample increased their triaxility with MHD physics. 
However, this small fraction correspond to halos that already were
outliers in the DMO sample and had triaxilities around $1/3$,
considerably lower than its parent sample.

@@

Figure \ref{fig:alignment_correlations} shows the correlation between
twisting at the four different radii at which we have measured the
alignments. 
The median, mean and standard deviation of the twisting are $2$, $6$ and $8$ degrees, respectively.
The correlation between twisting and triaxility also changes with
radius as measured by the Spearman's rank correlation coefficient.
The largest correlation shows with the halo triaxility measured at
$0.12R_{200}$. 
The most important consequence of this changing degree in correlation
is that twisting cannot only be explained by large triaxiality values
and a possibly noisy determination of the shape axis.

@@

Figure \ref{fig:disc_correlations} shows the correlations of
these quantities with the triaxility at five different radii.
We use the Spearmen's rank correlation coefficient to quantify the
correlation strength.
We find that the strongest correlations are found with the halo shape
measured at radii smaller than $0.12R_{200}$.
The  trend is such that halos with large triaxiality correlate with
high gas density and stellar discs with low mass and small size. 
In turn, massive and large stellar discs within with low density of
gas in its core correlate with low halo triaxiliaty.

In some cases, the halo shape also twists in the presence of baryons, as opposed to
the situation in DMO simulations where the shape is coherent as a
function of radius.
In Figure \ref{fig:twisting_correlations} we show the correlation
between twisting and the same baryonic properties that showed a high
degree of correlation with the triaxility.
In this case the disc to total ratio does not have a significant
degree of correlation.
Instead, the disc optical radius and the gas density at the core are
the most correlated. 
The correlation is such that extended stellar discs with a very low
gas density at its core show the most coherent halo shapes.

{\bf Recently \cite{2017MNRAS.469..594B} reported that MW-like DM haloes can
be roughly splint into two distinct families: stalled and accreting.
Stalled haloes assembled early and have remained almost unchanged
in the last $7$ Gyr, while accreting halos have been growing
constantly.
\cite{2017MNRAS.469..594B} also link the two populations to
differences in the dark matter filaments feeding the halos. 
Stalled halos at $z=0$ are embbeded in a single large thick filament, while
accreting halos are pierced by many thinner filaments. 
To what extent the twisting and high triaxility present in the MHD
halos can be linked to this dicotomy?
In Figure \ref{fig:twisting_age} we show the correlation between the
twisting and the age of the bulge, the age of the stellar disc and the
DM halo formation redshift. 
The ages are estimated directly as the mean value in the stellar age
distribution for each component, while the formation redshift is
defined as the redshif at which the halo already has half of its $z=0$
mass. 
In the three cases we find a correlation between ages and twisting
that goes in the direction claimed by \cite{2017MNRAS.469..594B}:
components with early assembly (i.e. stalled) have a smaller degree of
twisting.
A detailed study with respect to the cosmic web location is beyond the
scope of our paper.}

@@

We now use the results reported by \cite{LM10} and \cite{Bovy16}
to place our results in an observational context.
\cite{LM10} used observations of the Sagittarius tidal stream to
constrain the shape of the gravitational potential.
Their point of depart is that previous studies that assumed an
axisymetric galactic potential were not able to fit all the available
dynamic constraints for the Sagittarius stream, therefore making
necessary the use of a rigid triaxial potential with coaxial potential
ellipsoids for the dark matter component.  
Their results constrain the triaxility of this potential
component. 
They also translate their results into a triaxiality of the density
contours (that could be compared against our results)
 to be $(c/a)=0.44$ and $(b/a)=0.97$ at a radius of $\sim 40$kpc. 
They do not report any uncertainties for these two values. 
Looking at their plots for the quality of fit criterion as a function
of dark halo axial scales (their Figure 5), we choose a conservative $10\%$
relative uncertainty.
One surprising element in their results is that  the major axis of the
halo shape is perpendicular to the stellar disc plane.

@@

Figure \ref{fig:observations} shows an explicit comparison in
the $c/a$-$b/a$ plane of our results in MHD simulations against the
results by \cite{LM10} and \cite{Bovy16}.
We find six MHD halos with $b/a<0.93$ and $c/a>0.77$ that could be
considered consistent with their shape constraints  by \cite{Bovy16}, while
only one outlier DMO halo is consistent with those constraints.
In contrast, none of the simulated halos (MHD nor DMO) is consistent
with the results \cite{LM10}. 
The change of triaxility with radius in our simulations cannot account
for these two extremely different shape constraints at different
radii.

@@

We find that MHD halos are rounder than DMO halos at every radius.
MHD halos tend towards more oblate, sometimes almost spherical, shapes
($T<1/3$), while DMO halos tend towards more prolate shapes ($T>2/3$).  
The rounding effect is more noticeable as one moves closer to the galactic
disc and strongly  correlates with baryonic properties of the disc
More precisely, the triaxility is smaller for extended and massive
stellar discs with low gas densities at its core.

>>>>

\begin{figure*}
\begin{center}
\includegraphics[width=0.8\columnwidth]{Lvl_4_Triax_Plane_DM.pdf}
\includegraphics[width=0.8\columnwidth]{Lvl_4_Triax_Plane_MHD.pdf}
\end{center}
\caption{Axial ratios for all simulated halos.
  Left/right panel corresponds to DMO/MHD simulations, respectively.
  Triangles (squares) represent the measurements at $R_{200}/16$
  ($R_{200}$) that correspond to physical distances of $\approx14$ kpc
  ($\approx 230$ kpc).
  Here we visualize three main population trends.
  First, in DMO simulations halos are rounder in the outskirts
  than in the inner part.
  Second, halos in MHD are rounder in the inner regions than in
  the outskirts (opposite to the DMO trend). 
  Third, halos in MHD are rounder than its DMO counterparts.
  These three points are further clarified in Figure 
  The first and second points are claryfied in Figure
  \ref{fig:triaxial_cumulative} through cumulative triaxiality
  distributions}
  \label{fig:triaxiality_plane}
\end{figure*}

\begin{figure*}
\includegraphics[width=0.32\textwidth]{triaxialiy_distro_DM.pdf}
\includegraphics[width=0.32\textwidth]{triaxialiy_distro_MHD.pdf}
\includegraphics[width=0.32\textwidth]{delta_triaxiality_distro.pdf}
\caption{Triaxiality cumulative distribution at five different radii.
  Right/middle panel corresponds to the cumulative distribution of
  triaxiality in DMO/MHD simulations, respectively. 
  In DMO simulations the median triaxiality at all radii is larger
  than $2/3$. Furthermore, the triaxiality increases as one moves
  towards the inner part of the halo.
  In MHD simulations these trend reverses.
  The median triaxiality at all radii is smaller than $1/3$ and the
  halo is less triaxial as on moves towards the stellar disc.
  The right panel, shows the cumulative distribution for the changes
  in triaxiality in simulations with different physics, $\Delta
  T=T_{\rm MHD}-T_{\rm DMO}$. 
  At the virial radius all halos become less triaxial in the MHD
  simulations, the change is stronger towards the halo center.
  Only two haloes in the sample show the opposite trend of becoming
  more triaxial (but only at the smaller radii).} 
\label{fig:triaxial_cumulative}
\end{figure*}

@@

\begin{figure*}
\begin{center}
\includegraphics[width=1.0\textwidth]{correlations_twisting_triaxiality_MHD.pdf}
\end{center}
\caption{Twisting in the halo-disc alignment (measured as the standard deviation of the alignment angles)
  as a function of the halo triaxialiy at the radius indicated
  in each panel's title.
  The label with the $\rho$ value corresponds to the Spearmans rank
  correlation coefficient (mean value and uncertainty estimated via
  jackknife resampling) while the line shows the result of the
  minimum squares fit to the data.
  Twisting does not correlate with triaxiality in the same way at every
  radii. The strongest correlation appears around $0.12R_{200}$ ($28\pm2$ kpc).} 
\label{fig:alignment_correlations}
\end{figure*}

\begin{figure*}
\begin{center}
\includegraphics[width=0.8\textwidth]{correlation_T_MHD_disc.pdf}
\end{center}
\caption{Correlations between the halo triaxiality at different radii
  and baryonic disc properties. 
  The label with the $\rho$ value corresponds to the Spearmans rank
  correlation coefficient (mean value and uncertainty estimated via
  jackknife resampling).
  The line is the best minimum squares fit to a line.
  The x-axis in the first column is the gas density at the center of
  the galaxy with in a sphere of radius  $1$ kpc \citep{Pakmor17};
  the second column shows the disc to total mass ratio and the last
  column includes the disc optical radius defined to be the radius at which the
  $B$-band surface brightness drops below 25 mag arcsec$^{-2}$ \citep{auriga}.
  The largest correlations are found for the two smaller radii
  ($R_{200}/16$ and $R_{200}/8$).
  Extended and massive stellar discs with a low gas content at its
  core are correlated with low dark matter triaxilities.
  The correlation decreases as one approaches larger radii.}
\label{fig:disc_correlations}
\end{figure*}

\begin{figure*}
\begin{center}
\includegraphics[width=0.9\textwidth]{correlations_angles_alignment_MHD.pdf}
\end{center}
\caption{Twisting in the halo-disc alingment as a function of the same
  baryonic disc properties in Figure \ref{fig:disc_correlations}.
  The label with the $\rho$ value corresponds to the Spearmans rank
  correlation coefficient (mean value and uncertainty estimated via
  jackknife resampling).
  The line is the best minimum squares fit to a line.
  In this case the disc to total mass ratio does not correlate with the alignment twisting
  The gas density at the center and the disc optical radius still show
  a strong correlation, albeit weaker than what they had with the triaxiality.}
\label{fig:twisting_correlations}
\end{figure*}

@@

Figure \ref{fig:resolution} shows the results for the six halos
available both at Level3 and Level4 resolutions.
The dotted lines compare the DMO simulations showing that down to
scales of $1$ kpc the two resolutions yield comparable results
\citep{VeraCiro11}. 
In MHD simulations the situation is different, as expected. 
The shape ratios $b/a$ and $c/a$ start to differentiate
from each other at radii smaller than $15$ kpc, tith the exception of
Au-27 that shows a good agreement in the shape measuremnts at all radii, 
Above $15$ kpc, the differences between the triaxiality values at
Level4 and Level3 in MHD simulations are small and never exceed $0.1$. 
For this reason, in order to make the DM halo shape results comparable
in DMO and MHD simulations we only report measurements above $15$
kpc.}

@@

{\bf The main measurements are summarized in Table \ref{table:DMO} for the
DMO simulations and Table \ref{table:MHD} for the MHD simulations.
These results are computed at five different radii $R_{200}/16$,
$R_{200}/8$, $R_{200}/4$, $R_{200}/2$ and $R_{200}$.
In the following subsections we present and comment in greater detail
the results for the triaxiality, the halo-disc alignment and the
correlations with baryonic disc properties.}

\subsection{Triaxiality}

@@

In Figure \ref{fig:triaxial_cumulative} we summarize the results at
different radii using the cumulative distributions for the 
triaxiality parameter $T$ defined as

@@

The left panel of Figure \ref{fig:triaxial_cumulative} shows that in
the DMO sample the triaxiality has a median larger than $2/3$ (a
typical value that marks the transition from triaxiality to
prolateness) at all radii. 
Furthermore this median value increases as we move towards the inner part of the
halo.
The middle panel in the same Figure \ref{fig:triaxial_cumulative}
shows how the trends in the MHD simulations go 
in the opposite direction with respect to DMO results.
There the median triaxiality is always smaller than $1/3$ (a typical
values that marks the transition from triaxiality to prolateness) and this
median value decreases as we move towards the inner part of the halo.

We now quantify this change in triaxiality at the level of individual halos.
We compute $\Delta T\equiv T_{\rm MHD}-T_{\rm DMO}$ to quantify the
change between the triaxiality in the MHD and the DMO simulation.
The right panel in Figure \ref{fig:delta_triaxial_cumulative} shows
the cumulative distribution at the same radii as in the other panels.
There we observe that at $R_{200}$ all the halos have
reduced its triaxiality after including baryonic physics.
At smaller radii this general trend continues, although 2 haloes
(Au-11 and Au-17)  in our sample increased their triaxiality with MHD physics. 
However, this small fraction correspond to halos that already were
outliers in the DMO sample and had triaxilities around $1/3$,
considerably lower than its parent sample.

@@

Figure \ref{fig:alignment_correlations} shows the correlation between
twisting at the four different radii at which we have measured the
alignments. 
The median, mean and standard deviation of the twisting are $2$, $6$ and $8$ degrees, respectively.
The correlation between twisting and triaxiality also changes with
radius as measured by the Spearman's rank correlation coefficient.
The largest correlation shows with the halo triaxiality measured at
$0.12R_{200}$. 
The most important consequence of this changing degree in correlation
is that twisting cannot only be explained by large triaxiality values
and a possibly noisy determination of the shape axis.

@@

Figure \ref{fig:disc_correlations} shows the correlations of
these quantities with the triaxiality at five different radii.
We use the Spearmen's rank correlation coefficient to quantify the
correlation strength.
We find that the strongest correlations are found with the halo shape
measured at radii smaller than $0.12R_{200}$.
The  trend is such that halos with large triaxiality correlate with
high gas density and stellar discs with low mass and small size. 
In turn, massive and large stellar discs within with low density of
gas in its core correlate with low halo triaxiliaty.

In some cases, the halo shape also twists in the presence of baryons, as opposed to
the situation in DMO simulations where the shape is coherent as a
function of radius.
In Figure \ref{fig:twisting_correlations} we show the correlation
between twisting and the same baryonic properties that showed a high
degree of correlation with the triaxiality.
In this case the disc to total ratio does not have a significant
degree of correlation.
Instead, the disc optical radius and the gas density at the core are
the most correlated. 
The correlation is such that extended stellar discs with a very low
gas density at its core show the most coherent halo shapes.

{\bf Recently \cite{2017MNRAS.469..594B} reported that MW-like DM haloes can
be roughly splint into two distinct families: stalled and accreting.
Stalled haloes assembled early and have remained almost unchanged
in the last $7$ Gyr, while accreting halos have been growing
constantly.
\cite{2017MNRAS.469..594B} also link the two populations to
differences in the dark matter filaments feeding the halos. 
Stalled halos at $z=0$ are embbeded in a single large thick filament, while
accreting halos are pierced by many thinner filaments. 
To what extent the twisting and high triaxiality present in the MHD
halos can be linked to this dicotomy?
In Figure \ref{fig:twisting_age} we show the correlation between the
twisting and the age of the bulge, the age of the stellar disc and the
DM halo formation redshift. 
The ages are estimated directly as the mean value in the stellar age
distribution for each component, while the formation redshift is
defined as the redshif at which the halo already has half of its $z=0$
mass. 
In the three cases we find a correlation between ages and twisting
that goes in the direction claimed by \cite{2017MNRAS.469..594B}:
components with early assembly (i.e. stalled) have a smaller degree of
twisting.
A detailed study with respect to the cosmic web location is beyond the
scope of our paper.}

@@

We now use the results reported by \cite{LM10} and \cite{Bovy16}
to place our results in an observational context.
\cite{LM10} used observations of the Sagittarius tidal stream to
constrain the shape of the gravitational potential.
Their point of depart is that previous studies that assumed an
axisymetric galactic potential were not able to fit all the available
dynamic constraints for the Sagittarius stream, therefore making
necessary the use of a rigid triaxial potential with coaxial potential
ellipsoids for the dark matter component.  
Their results constrain the triaxiality of this potential
component. 
They also translate their results into a triaxiality of the density
contours (that could be compared against our results)
 to be $(c/a)=0.44$ and $(b/a)=0.97$ at a radius of $\sim 40$kpc. 
They do not report any uncertainties for these two values. 
Looking at their plots for the quality of fit criterion as a function
of dark halo axial scales (their Figure 5), we choose a conservative $10\%$
relative uncertainty.
One surprising element in their results is that  the major axis of the
halo shape is perpendicular to the stellar disc plane.

@@

Figure \ref{fig:observations} shows an explicit comparison in
the $c/a$-$b/a$ plane of our results in MHD simulations against the
results by \cite{LM10} and \cite{Bovy16}.
We find six MHD halos with $b/a<0.93$ and $c/a>0.77$ that could be
considered consistent with their shape constraints  by \cite{Bovy16}, while
only one outlier DMO halo is consistent with those constraints.
In contrast, none of the simulated halos (MHD nor DMO) is consistent
with the results \cite{LM10}. 
The change of triaxiality with radius in our simulations cannot account
for these two extremely different shape constraints at different
radii.

@@

We find that MHD halos are rounder than DMO halos at every radius.
MHD halos tend towards more oblate, sometimes almost spherical, shapes
($T<1/3$), while DMO halos tend towards more prolate shapes ($T>2/3$).  
The rounding effect is more noticeable as one moves closer to the galactic
disc and strongly  correlates with baryonic properties of the disc
More precisely, the triaxiality is smaller for extended and massive
stellar discs with low gas densities at its core.
